---
title: "Data Cleaning"
output: html_document
date: "2023-12-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(plotly)
library(shiny)
library(shinydashboard)
library(cowplot)
```


## Import Dataset

```{r import}

est_cases = read.csv("Est Mal cases by country.csv", sep = ",",header=TRUE, skip = 1)
conf_cases = read.csv("MALARIA_CONFIRMED_CASES.csv", sep = ",", header = TRUE, skip = 1)
pres_cases = read.csv("MALARIA_PRESUMED_CASES.csv", sep = ",", header = TRUE, skip = 1)
tot_cases = read.csv("MALARIA_TOTAL_CASES_Presumed_confirmed.csv",sep = ",", header = TRUE, skip = 1)
est_incd = read.csv("MALARIA_EST_INCIDENCE_per1000.csv", sep = ",", header = TRUE, skip = 1)
imported_cases = read.csv("MALARIA_IMPORTED.csv", sep = ",", header = TRUE, skip = 1)
ing_cases = read.csv("MALARIA_INDIG.csv", sep = ",", header = TRUE)

#Death
est_death = read.csv("MALARIA_EST_DEATHS.csv", sep = ",", header = TRUE, skip = 1)
est_mort = read.csv("MALARIA_EST_MORTALITY.csv", sep = ",", header = TRUE, skip = 1)

world_countries_by_region = read.csv("world_countries_by_region.csv", sep = ",", header = TRUE)

```


## Clean Data

1. Estimated Cases 

```{r est}
#Convert name of headers
colnames(est_cases) <- c("Country", sub("^X", "", colnames(est_cases)[-1]))


est_cases_long <- est_cases %>%
  pivot_longer(
    cols = -Country,  
    names_to = "Year",  
    values_to = "Cases" 
  )

est_cases_long <- est_cases_long %>%
  separate(
    col = Cases,  # Column to split
    into = c("Est_Cases", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


est_cases_long <- est_cases_long %>%
  mutate(
    Est_Cases = as.numeric(gsub("[^0-9.]", "", Est_Cases)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(est_cases_long)
head(est_cases_long)


#Summary Statistics
summary_stats <- est_cases_long %>%
  group_by(Country) %>%
  summarise(
    Average_Cases = mean(Est_Cases, na.rm = TRUE),
    Total_Cases = sum(Est_Cases, na.rm = TRUE),
    Min_Cases = min(Est_Cases, na.rm = TRUE),
    Max_Cases = max(Est_Cases, na.rm = TRUE)
  )
summary_stats


```


## Match countries


```{r}

country_name_corrections <- c(
  "Bolivia (Plurinational State of)" = "Bolivia",
  "Cote d'Ivoire" = "Côte d'Ivoire",
  "Cabo Verde" = "Cape Verde",
  "Democratic People's Republic of Korea" = "Dem. Rep. Korea",
  "Democratic Republic of the Congo" = "Dem. Rep. Congo",
  "Equatorial Guinea" = "Eq. Guinea", # Check if map data uses a different name
  "Eswatini" = "Swaziland",
  "Iran (Islamic Republic of)" = "Iran",
  "Lao People's Democratic Republic" = "Lao PDR",
  "Republic of Korea" = "Korea",
  "Sao Tome and Principe" = "São Tomé and Principe", # Check the exact spelling in the map data
  "Solomon Islands" = "Solomon Is.", # Check if map data uses a different name
  "South Sudan" = "S. Sudan", # Check if map data uses a different name
  "Syrian Arab Republic" = "Syria",
  "Turkiye" = "Turkey",
  "United Republic of Tanzania" = "Tanzania",
  "Venezuela (Bolivarian Republic of)" = "Venezuela",
  "Viet Nam" = "Vietnam",
  "Central African Republic" = "Central African Rep.",
  "Dominican Republic" = "Dominican Rep.")

est_cases_long$Country <- ifelse(est_cases_long$Country %in% names(country_name_corrections),       country_name_corrections[est_cases_long$Country],
                                 est_cases_long$Country)

```

## 2. Confirmed Cases

```{r}
colnames(conf_cases) = c("Country", sub("^.", "", colnames(conf_cases)[-1]))
conf_cases <- conf_cases %>%
  mutate(across(-Country, ~ as.numeric(gsub("[ ,]", "", .))))
#head(conf_cases)

conf_cases_long <- conf_cases %>%
  pivot_longer(
    cols = -Country,  # Select all columns except for Country to pivot
    names_to = "Year",  # This will create the new 'Year' header
    values_to = "Cases"  # This will contain the values for each year
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert Year to numeric

head(conf_cases_long)
```



3. Presumed Cases

```{r}
colnames(pres_cases) = c("Country", sub("^X.", "", colnames(pres_cases)[-1]))
pres_cases <- pres_cases %>%
  mutate(across(-Country, ~ as.numeric(gsub("[ ,]", "", .))))
#head(pres_cases)

pres_cases_long <- pres_cases %>%
  pivot_longer(
    cols = -Country,  # Select all columns except for Country to pivot
    names_to = "Year",  # This will create the new 'Year' header
    values_to = "Cases"  # This will contain the values for each year
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert Year to numeric

head(pres_cases_long)
```


4. Total Cased: Presumed + Confirmed

```{r}
head(tot_cases)
colnames(tot_cases) = c("Country", sub("^X.", "", colnames(tot_cases)[-1]))
tot_cases <- tot_cases %>%
  mutate(across(-Country, ~ as.numeric(gsub("[ ,]", "", .))))
#head(tot_cases)

tot_cases_long <- tot_cases %>%
  pivot_longer(
    cols = -Country,  # Select all columns except for Country to pivot
    names_to = "Year",  # This will create the new 'Year' header
    values_to = "Cases"  # This will contain the values for each year
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert Year to numeric

head(tot_cases_long)

```

5. Estimated Incidence per 1000

```{r}
#Convert name of headers
head(est_incd)
colnames(est_incd) <- c("Country", sub("^X.", "", colnames(est_incd)[-1]))


est_incd_long <- est_incd %>%
  pivot_longer(
    cols = -Country,  
    names_to = "Year",  
    values_to = "Cases" 
  )

est_incd_long <- est_incd_long %>%
  separate(
    col = Cases,  # Column to split
    into = c("Est_Incd", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


est_incd_long <- est_incd_long %>%
  mutate(
    Est_Incd = as.numeric(gsub("[^0-9.]", "", Est_Incd)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(est_cases_long)
head(est_incd_long)
```


6. Imported Cases

```{r}
colnames(imported_cases) = c("Country", sub("^X.", "", colnames(imported_cases)[-1]))
imported_cases <- imported_cases %>%
  mutate(across(-Country, ~ as.numeric(gsub("[ ,]", "", .))))
#head(conf_cases)

imported_cases_long <- imported_cases %>%
  pivot_longer(
    cols = -Country,  # Select all columns except for Country to pivot
    names_to = "Year",  # This will create the new 'Year' header
    values_to = "Cases"  # This will contain the values for each year
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert Year to numeric

head(imported_cases_long)
```


7. Indigeneous Cases

```{r}
colnames(ing_cases) = c("Country", sub("^X.", "", colnames(ing_cases)[-1]))
ing_cases<- ing_cases %>%
  mutate(across(-Country, ~ as.numeric(gsub("[ ,]", "", .))))
#head(conf_cases)

ing_cases_long <- ing_cases %>%
  pivot_longer(
    cols = -Country,  # Select all columns except for Country to pivot
    names_to = "Year",  # This will create the new 'Year' header
    values_to = "Cases"  # This will contain the values for each year
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert Year to numeric

head(ing_cases_long)
```


8. Estimated Death
```{r}
#Convert name of headers
head(est_death)
colnames(est_death) <- c("Country", sub("^X.", "", colnames(est_death)[-1]))


est_death_long <-est_death %>%
  pivot_longer(
    cols = -Country,  
    names_to = "Year",  
    values_to = "Cases" 
  )

est_death_long <- est_death_long %>%
  separate(
    col = Cases,  # Column to split
    into = c("Est_Death", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


est_death_long <- est_death_long %>%
  mutate(
    Est_Death = as.numeric(gsub("[^0-9.]", "", Est_Death)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(est_cases_long)
head(est_death_long)
```


9. Estimated Mortality Rates
```{r}
#Convert name of headers
#head(est_mort)
colnames(est_mort) <- c("Country", sub("^X.", "", colnames(est_mort)[-1]))


est_mort_long <-est_mort %>%
  pivot_longer(
    cols = -Country,  
    names_to = "Year",  
    values_to = "Cases" 
  )

est_mort_long <- est_mort_long %>%
  separate(
    col = Cases,  # Column to split
    into = c("Est_Mort", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


est_mort_long <- est_mort_long %>%
  mutate(
   Est_Mort = as.numeric(gsub("[^0-9.]", "", Est_Mort)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(est_cases_long)
head(est_mort_long)

```

Save clean dataset to a new CSV file

```{r}
# Create a named list of dataframes
#dataframes <- list(
#   est_cases_long = est_cases_long,
#   conf_cases_long = conf_cases_long,
#   pres_cases_long = pres_cases_long,
#   tot_cases_long = tot_cases_long,
#   est_incd_long = est_incd_long,
#   imported_cases_long = imported_cases_long,
#   ing_cases_long = ing_cases_long,
#   est_death_long = est_death_long,
#   est_mort_long = est_mort_long
# )

# Iterate over the list and write each dataframe to a CSV file
#lapply(names(dataframes), function(name) {
  #write.csv(dataframes[[name]], paste0(name, ".csv"), row.names = FALSE)
#})

```


## Global Cases
 
## Global Cases

```{r}
global_est_cases = read.csv("MALARIA_EST_CASES_Global.csv", sep = ",",header=TRUE, skip = 1)
global_conf_cases = read.csv("MALARIA_CONF_CASES_Global.csv", sep = ",", header = TRUE, skip = 1)
global_pres_cases = read.csv("MALARIA_PRES_CASES_Global.csv", sep = ",", header = TRUE, skip = 1)
global_tot_cases = read.csv("MALARIA_TOTAL_CASES_Global.csv",sep = ",", header = TRUE, skip = 1)
global_est_incd = read.csv("MALARIA_EST_INCIDENCE_Global.csv", sep = ",", header = TRUE, skip = 1)
global_imported_cases = read.csv("MALARIA_IMPORTED_Global.csv", sep = ",", header = TRUE, skip = 1)
global_ing_cases = read.csv("MALARIA_INDIG_Global.csv", sep = ",", header = TRUE)


#Death
global_est_death = read.csv("MALARIA_EST_DEATHS_Global.csv", sep = ",", header = TRUE, skip = 1)
global_est_mort = read.csv("MALARIA_EST_MORTALITY_Global.csv", sep = ",", header = TRUE, skip = 1)





```


## Global Data Cleaning

1. Global Estimated Cases
```{r}
#Convert name of headers
colnames(global_est_cases) <- c("Region", sub("^X.", "", colnames(global_est_cases)[-1]))


global_est_cases_long <- global_est_cases %>%
  pivot_longer(
    cols = -Region,  
    names_to = "Year",  
    values_to = "Cases" 
  )

global_est_cases_long <- global_est_cases_long %>%
  separate(
    col = Cases,  # Column to split
    into = c("Global_Est_Cases", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


global_est_cases_long <- global_est_cases_long %>%
  mutate(
    Global_Est_Cases = as.numeric(gsub("[^0-9.]", "", Global_Est_Cases)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(est_cases_long)
head(global_est_cases_long)


#Summary Statistics
summary_stats <- global_est_cases_long %>%
  group_by(Region) %>%
  summarise(
    Average_Cases = mean(Global_Est_Cases, na.rm = TRUE),
    Total_Cases = sum(Global_Est_Cases, na.rm = TRUE),
    Min_Cases = min(Global_Est_Cases, na.rm = TRUE),
    Max_Cases = max(Global_Est_Cases, na.rm = TRUE)
  )
summary_stats
```

2. Global Confirmed Cases

```{r}
colnames(global_conf_cases) = c("Region", sub("^X.", "", colnames(global_conf_cases)[-1]))
global_conf_cases <- global_conf_cases %>%
  mutate(across(-Region, ~ as.numeric(gsub("[ ,]", "", .))))
#head(conf_cases)

global_conf_cases_long <- global_conf_cases %>%
  pivot_longer(
    cols = -Region,  # Select all columns except for Country to pivot
    names_to = "Year",  # This will create the new 'Year' header
    values_to = "Cases"  # This will contain the values for each year
  ) %>%
  mutate(Year = as.numeric(Year))  # Convert Year to numeric

head(global_conf_cases_long)
```



3. Global Death

```{r}
#Convert name of headers
head(global_est_death)
colnames(global_est_death) <- c("Region", sub("^X.", "", colnames(global_est_death)[-1]))


global_est_death_long <-global_est_death %>%
  pivot_longer(
    cols = -Region,  
    names_to = "Year",  
    values_to = "Cases" 
  )

global_est_death_long <- global_est_death_long %>%
  separate(
    col = Cases,  # Column to split
    into = c("Global_Est_Death", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


global_est_death_long <- global_est_death_long %>%
  mutate(
    Global_Est_Death = as.numeric(gsub("[^0-9.]", "", Global_Est_Death)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(global_est_cases_long)
head(global_est_death_long)
```


4. Global Mortality

```{r}
#Convert name of headers
#head(est_mort)
colnames(global_est_mort) <- c("Region", sub("^X.", "", colnames(global_est_mort)[-1]))


global_est_mort_long <-global_est_mort %>%
  pivot_longer(
    cols = -Region,  
    names_to = "Year",  
    values_to = "Global_Cases" 
  )

global_est_mort_long <- global_est_mort_long %>%
  separate(
    col = Global_Cases,  # Column to split
    into = c("Global_Est_Mort", "Range"),  # New columns to create
    sep = "\\s+(?=\\[)",  # Regular expression to split by space before '['
    remove = FALSE,  # Remove the original column if you don't need it anymore
    convert = FALSE  # Set to FALSE to keep everything as character strings for now
  )


global_est_mort_long <- global_est_mort_long %>%
  mutate(
   Global_Est_Mort = as.numeric(gsub("[^0-9.]", "", Global_Est_Mort)),  # Clean and convert to numeric
    Range = gsub("\\[|\\]", "", Range),  # Remove brackets
    Year = as.numeric(Year)  # Convert Year to numeric
  )
#str(est_cases_long)
head(global_est_mort_long)


# dataframes <- list(
#    global_est_cases_long = global_est_cases_long,
#    global_conf_cases_long = global_conf_cases_long,
#    global_est_death_long = global_est_death_long,
#    global_est_mort_long = global_est_mort_long
# )
# 
# # Iterate over the list and write each dataframe to a CSV file
# lapply(names(dataframes), function(name) {
# write.csv(dataframes[[name]], paste0(name, ".csv"), row.names = FALSE)
# })
```




Bubble Plot for Estimated Deaths

```{r}
# plot <- est_death_long %>%
#   left_join(est_mort_long, by = c("Country", "Year")) %>%
#   plot_ly(
#     x = ~Year, 
#     y = ~Est_Death, 
#     text = ~Country, 
#     size = ~Est_Mort, 
#     sizes = c(20, 60), 
#     color = ~Country,
#     colors = "RdYlBu"
#   ) %>%
#   add_markers() %>%
#   layout(title = 'Estimated Malaria Deaths and Mortality Rates by Year and Country',
#          xaxis = list(title = 'Year'),
#          yaxis = list(title = 'Estimated Deaths'))
# 
# # To view the plot, simply run this line in your R environment
# plot
```