---
title: "Cleaning and Visualization"
output: html_document
date: "2023-12-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(sf)
library(rnaturalearth)
library(plotly)
library(shiny)
library(shinydashboard)
library(cowplot)
library(treemap)
library(readr)
library(stringr)
library(forecast)
library(caTools)
library(caret)
```

## Import Data

```{r import}
world_countries_by_region = read.csv("world_countries_by_region.csv", sep = ",", header = TRUE)
est_cases_long = read.csv("est_cases_long.csv", sep = ",", header = TRUE)
conf_cases_long = read.csv("conf_cases_long.csv", sep = ",", header = TRUE)
pres_cases_long = read.csv("pres_cases_long.csv", sep = ",", header = TRUE)
tot_cases_long = read.csv("tot_cases_long.csv", sep = ",", header = TRUE)
est_incd_long = read.csv("est_incd_long.csv", sep = ",", header = TRUE)
imported_cases_long = read.csv("imported_cases_long.csv", sep = ",", header = TRUE)
ing_cases_long = read.csv("ing_cases_long.csv", sep = ",", header = TRUE)


est_death_long = read.csv("est_death_long.csv", sep = ",", header = TRUE)
est_mort_long = read.csv("est_mort_long.csv", sep = ",", header = TRUE)


global_est_cases_long = read.csv("global_est_cases_long.csv", sep = ",", header = TRUE)
global_conf_cases_long = read.csv("global_conf_cases_long.csv", sep = ",", header = TRUE)
global_est_death_long = read.csv("global_est_death_long.csv", sep = ",", header = TRUE)
global_est_mort_long = read.csv("global_est_mort_long.csv", sep = ",", header = TRUE)
```




## Merge Data

```{r merge}
merged_death = read.csv("merged_death.csv", sep = ",", header = TRUE)
merged_mort = read.csv("merged_mort.csv", sep = ",", header = TRUE)
#$Est_Mort <- as.numeric(merged_mort$Est_Mort)
merged_est = read.csv("merged_est.csv", sep = ",", header = TRUE)
merged_tot = read.csv("merged_tot.csv", sep = ",", header = TRUE)
merged_imp = read.csv("merged_imp.csv", sep = ",", header = TRUE)
merged_ing = read.csv("merged_ing.csv", sep = ",", header = TRUE)
merged_incd = read.csv("merged_incd.csv", sep = ",", header = TRUE)

merged_deathh <- select(merged_death, -Cases, -Range)
merged_incdd <- select(merged_incd, -Cases, -Range)
merged_estt <- select(merged_est, -Cases, -Range)


merged_data <- merge(merged_deathh, merged_incdd, by=c("Year","Country","Region"))
merged_data <- merge(merged_data, merged_estt,  by=c("Year","Country","Region"))
attach(merged_data)

merged_use = select(merged_data, -Country, - Region)

```


## Adjusting data country names based on R country names

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf")

country_name_corrections <- c(
  "Bolivia (Plurinational State of)" = "Bolivia",
  "Cote d'Ivoire" = "Côte d'Ivoire",
  "Cabo Verde" = "Cape Verde",
  "Democratic People's Republic of Korea" = "Dem. Rep. Korea",
  "Democratic Republic of the Congo" = "Dem. Rep. Congo",
  "Equatorial Guinea" = "Eq. Guinea", # Check if map data uses a different name
  "Eswatini" = "Swaziland",
  "Iran (Islamic Republic of)" = "Iran",
  "Lao People's Democratic Republic" = "Lao PDR",
  "Republic of Korea" = "Korea",
  "Sao Tome and Principe" = "São Tomé and Principe", # Check the exact spelling in the map data
  "Solomon Islands" = "Solomon Is.", # Check if map data uses a different name
  "South Sudan" = "S. Sudan", # Check if map data uses a different name
  "Syrian Arab Republic" = "Syria",
  "Turkiye" = "Turkey",
  "United Republic of Tanzania" = "Tanzania",
  "Venezuela (Bolivarian Republic of)" = "Venezuela",
  "Viet Nam" = "Vietnam",
  "Central African Republic" = "Central African Rep.",
  "Dominican Republic" = "Dominican Rep.")


est_cases_long$Country <- ifelse(est_cases_long$Country %in% names(country_name_corrections),       country_name_corrections[est_cases_long$Country],
                                 est_cases_long$Country)




mismatched_countries <- setdiff(est_cases_long$Country, world$name)
#Two countries were mismatched (this means the world package does not have these countries named correctly)
#"French Guiana" "Mayotte"

mismatched_countries

```


## Malaria trend world wide visualization for the year 2000

```{r trend_2000}
#year 2000 
data_2000 <- est_cases_long %>%
  filter(Year == 2000) %>%
  select(Country, Est_Cases)

# Merge the map data with the 2000 cases data
world_data_2000 <- world %>%
  left_join(data_2000, by = c("name" = "Country"))

# Plot the map
ggplot(data = world_data_2000) +
  geom_sf(aes(fill = Est_Cases), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey", 
                       labels = scales::comma, guide = guide_legend(title = "Estimated\nCases")) +
  labs(title = "Estimated Malaria Cases in 2000", 
       fill = "Cases") +
  theme_void() +
  theme(legend.position = "bottom")
```



## Malaria trend world wide visualization for the year 2010


```{r trend_2010}
#year 2010 
data_2010 <- est_cases_long %>%
  filter(Year == 2010) %>%
  select(Country, Est_Cases)

# Merge the map data with the 2010 cases data
world_data_2010 <- world %>%
  left_join(data_2010, by = c("name" = "Country"))

# Plot the map
ggplot(data = world_data_2010) +
  geom_sf(aes(fill = Est_Cases), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey", 
                       labels = scales::comma, guide = guide_legend(title = "Estimated\nCases")) +
  labs(title = "Estimated Malaria Cases in 2010", 
       fill = "Cases") +
  theme_void() +
  theme(legend.position = "bottom")
```



## Malaria trend world wide visualization for the year 2021

```{r trend_2021}
#year 2021 
data_2021 <- est_cases_long %>%
  filter(Year == 2021) %>%
  select(Country, Est_Cases)

# Merge the map data with the 2021 cases data
world_data_2021 <- world %>%
  left_join(data_2021, by = c("name" = "Country"))

# Plot the map
ggplot(data = world_data_2021) +
  geom_sf(aes(fill = Est_Cases), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey", 
                       labels = scales::comma, guide = guide_legend(title = "Estimated\nCases")) +
  labs(title = "Estimated Malaria Cases in 2021", 
       fill = "Cases") +
  theme_void() +
  theme(legend.position = "bottom")
```


## Total Confirmed Malaria Cases Over the Years

```{r confirmed}
annual_confirmed <- conf_cases_long %>%
  group_by(Year) %>%
  summarise(Total_Confirmed = sum(Cases, na.rm = TRUE))

# Create a time series plot
ggplot(annual_confirmed, aes(x = Year, y = Total_Confirmed)) +
  geom_line(color = "blue", size = 1) +
  theme_classic() +
  labs(title = "Total Confirmed Malaria Cases Over the Years",
       x = "Year",
       y = "Total Confirmed Cases")
```


## Aggregate death estimates for each year

```{r death}
 # Aggregate death estimates for each year
annual_deaths <- est_death_long %>%
  group_by(Year) %>%
  summarise(Total_Deaths = sum(Est_Death, na.rm = TRUE))

# Create a time series plot
ggplot(annual_deaths, aes(x = Year, y = Total_Deaths)) +
  geom_line(color = "blue", size = 1) +
  theme_classic() +
  labs(title = "Total Estimated Malaria Deaths Over the Years",
       x = "Year",
       y = "Total Estimated Deaths")
```




## Treemap of death cases

```{r}
# Sum the death estimates for each country
total_deaths_by_country <- est_death_long %>%
  group_by(Country) %>%
  summarise(Total_Deaths = sum(Est_Death, na.rm = TRUE)) %>%
  ungroup()

#Create a treemap
plot_ly(
  data = total_deaths_by_country,
  labels = ~Country,
  parents = NA,
  values = ~Total_Deaths,
  type = "treemap",
  textinfo = 'label+value',
  hoverinfo = 'label+value+percent parent'
) %>%
  layout(title = "Distribution of Total Estimated Deaths by Country")
```


## Bubble Plots

```{r bubble}

latest_year <- max(merged_death$Year)
merged_death_latest <- subset(merged_death, Year == latest_year)
merged_mort_latest <- subset(merged_mort, Year == latest_year)


# Merging the death and mortality data
merged_data <- merge(merged_death_latest, merged_mort_latest, by=c('Country', 'Year', 'Region'))


```

## Average death 

```{r avg}
# Sorting data to show highest death rates at the top
# Aggregating data by region (sum, mean, or median can be used based on requirement)
na_counts_per_column <- colSums(is.na(merged_death))
#print(na_counts_per_column)

region_data <- merged_death %>%
  group_by(Region) %>%
  summarise(Avg_Mortality = mean(Est_Death, na.rm = TRUE)) %>%
  arrange(desc(Avg_Mortality))

# Creating the bar chart for regions
ggplot(region_data, aes(x = Region, y = Avg_Mortality, fill = Region)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  labs(title = "Average Malaria Death Rates by Region",
       x = "Region",
       y = "Average Death Rate") +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "none")
```

##  

```{r mort_death}
# Creating a ggplot
ggplot(merged_data, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    scale_size(range=c(1, 10)) +
    labs(title='Malaria Estimates by Country and Region',
         x='Estimated Deaths in Latest Year',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right')





# p <- ggplot(merged_data, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
#     geom_point(alpha=0.7) +
#     geom_text(aes(label=Country), check_overlap = TRUE, vjust = "inward", hjust = "inward") +
#     scale_size(range=c(1, 10)) +
#     labs(title='Malaria Estimates by Country and Region (Interactive)',
#          x='Estimated Deaths in Latest Year',
#          y='Average Estimated Mortality Rate',
#          color='Region') +
#     theme_minimal() +
#     theme(legend.position='right')
# 
# # Converting to an interactive plotly plot
# ggplotly(p, tooltip = "text")


ggplot(merged_data, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    scale_size(range=c(1, 10)) +
    labs(title='Malaria Estimates by Country and Region',
         x='Estimated Deaths in Latest Year',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right')
```

## Death vs Mort for Americas

```{r america}

# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)


merged_death_americas <- subset(merged_death_2000, Region == 'Americas')
merged_mort_americas <- subset(merged_mort_2000, Region == 'Americas')
merged_incd_americas <- subset(merged_incd_2000, Region == 'Americas')

# Merging the death and mortality data for the Americas
merged_data_americas <- merge(merged_death_americas, merged_mort_americas, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the Americas region
p_americas <- ggplot(merged_data_americas, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
  scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in the Americas in 2000',
         x='Estimated Deaths in 2000',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_americas, tooltip = "text")





## 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)


merged_death_americas <- subset(merged_death_2021, Region == 'Americas')
merged_mort_americas <- subset(merged_mort_2021, Region == 'Americas')
merged_incd_americas <- subset(merged_incd_2021, Region == 'Americas')

# Merging the death and mortality data for the Americas
merged_data_americas <- merge(merged_death_americas, merged_mort_americas, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the Americas region
p_americas <- ggplot(merged_data_americas, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in the Americas in 2021',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_americas, tooltip = "text")

```





## Death vs Mort for Africans

```{r}

# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)

# Filtering data for the "Africa" region
merged_death_africa <- subset(merged_death_2000, Region == 'Africa')
merged_mort_africa <- subset(merged_mort_2000, Region == 'Africa')
merged_incd_africa <- subset(merged_incd_2000, Region == 'Africa')

# Merging the death and mortality data for the "Africa" region
merged_data_africa <- merge(merged_death_africa, merged_mort_africa, by=c('Country', 'Year', 'Region'))

p_africa <- ggplot(merged_data_africa, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1.5) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia"= "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in Africa in 2000',
         x='Estimated Deaths in Latest Year',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_africa, tooltip = "text")






## year 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)

# Filtering data for the "Africa" region
merged_death_africa <- subset(merged_death_2021, Region == 'Africa')
merged_mort_africa <- subset(merged_mort_2021, Region == 'Africa')
merged_incd_africa <- subset(merged_incd_2021, Region == 'Africa')

# Merging the death and mortality data for the "Africa" region
merged_data_africa <- merge(merged_death_africa, merged_mort_africa, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the Americas region
p_africa <- ggplot(merged_data_africa, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1.5) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in Africa in 2021',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right')

#Converting to an interactive plotly plot
ggplotly(p_africa, tooltip = "text")



```



## Death vs mort for South East Asia

```{r}

# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)


merged_death_asia <- subset(merged_death_2000, Region == 'South-East Asia')
merged_mort_asia <- subset(merged_mort_2000, Region == 'South-East Asia')
merged_incd_asia <- subset(merged_incd_2000, Region == 'South-East Asia')

# Merging the death and mortality data for the Asia
merged_data_asia <- merge(merged_death_asia, merged_mort_asia, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the Americas region
p_asia <- ggplot(merged_data_asia, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 2) +
    scale_size(range=c(1, 10)) +
  scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in South-East Asia in 2000',
         x='Estimated Deaths in 2000',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right') 

# Converting to an interactive plotly plot
ggplotly(p_asia, tooltip = "text")





## 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)


merged_death_asia <- subset(merged_death_2021, Region == 'South-East Asia')
merged_mort_asia <- subset(merged_mort_2021, Region == 'South-East Asia')
merged_incd_asia <- subset(merged_incd_2021, Region == 'South-East Asia')

# Merging the death and mortality data for the Americas
merged_data_asia <- merge(merged_death_asia, merged_mort_asia, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the Asia region
p_asia <- ggplot(merged_data_asia, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 2) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in South-East Asia in 2021',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_asia, tooltip = "text")

```




## Death vs mort for Eastern Mediterranean

```{r}
# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)


merged_death_em <- subset(merged_death_2000, Region == 'Eastern Mediterranean')
merged_mort_em <- subset(merged_mort_2000, Region == 'Eastern Mediterranean')
merged_incd_em <- subset(merged_incd_2000, Region == 'Eastern Mediterranean')

# Merging the death and mortality data for the E-Ms
merged_data_em <- merge(merged_death_em, merged_mort_em, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the E-Ms region
p_em <- ggplot(merged_data_em, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 2) +
    scale_size(range=c(1, 10)) +
  scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in Eastern Mediterranean in 2000',
         x='Estimated Deaths in 2000',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right') 

# Converting to an interactive plotly plot
ggplotly(p_em, tooltip = "text")





## 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)


merged_death_em <- subset(merged_death_2021, Region == 'Eastern Mediterranean')
merged_mort_em <- subset(merged_mort_2021, Region == 'Eastern Mediterranean')
merged_incd_em <- subset(merged_incd_2021, Region == 'Eastern Mediterranean')

# Merging the death and mortality data for the Americas
merged_data_em <- merge(merged_death_em, merged_mort_em, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the Americas region
p_em <- ggplot(merged_data_em, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 2) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in Eastern Mediterranean in 2021',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_em, tooltip = "text")


```


## Death vs mort for Western Pacific
```{r}
# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)


merged_death_wp <- subset(merged_death_2000, Region == 'Western Pacific')
merged_mort_wp <- subset(merged_mort_2000, Region == 'Western Pacific')
merged_incd_wp <- subset(merged_incd_2000, Region == 'Western Pacific')

# Merging the death and mortality data for the WPs
merged_data_wp <- merge(merged_death_wp, merged_mort_wp, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the WP region
p_wp <- ggplot(merged_data_wp, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
  scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in Western Pacific in 2000',
         x='Estimated Deaths in 2000',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right') 

# Converting to an interactive plotly plot
ggplotly(p_wp, tooltip = "text")





## 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)


merged_death_wp <- subset(merged_death_2021, Region == 'Western Pacific')
merged_mort_wp <- subset(merged_mort_2021, Region == 'Western Pacific')
merged_incd_wp <- subset(merged_incd_2021, Region == 'Western Pacific')

# Merging the death and mortality data for the Americas
merged_data_wp <- merge(merged_death_wp, merged_mort_wp, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the WP region
p_wp <- ggplot(merged_data_wp, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Death Estimates vs Malaria Mortality in Western Pacific in 2021',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_minimal() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_wp, tooltip = "text")
```


## Death vs mort for Europe

```{r}
# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)


merged_death_europe <- subset(merged_death_2000, Region == 'Europe')
merged_mort_europe <- subset(merged_mort_2000, Region == 'Europe')
merged_incd_europe <- subset(merged_incd_2000, Region == 'Europe')

# Merging the death and mortality data for the WPs
merged_data_europe <- merge(merged_death_europe, merged_mort_europe, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the WP region
p_europe <- ggplot(merged_data_europe, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
  scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "purple", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Europe Estimates (2000)',
         x='Estimated Deaths in 2000',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right') 

# Converting to an interactive plotly plot
ggplotly(p_europe, tooltip = "text")





## 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)


merged_death_europe <- subset(merged_death_2021, Region == 'Europe')
merged_mort_europe <- subset(merged_mort_2021, Region == 'Europe')
merged_incd_europe <- subset(merged_incd_2021, Region == 'Europe')

# Merging the death and mortality data for the Americas
merged_data_europe <- merge(merged_death_europe, merged_mort_europe, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the WP region
p_europe <- ggplot(merged_data_europe, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "purple", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='Europe Estimates (2021)',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_europe, tooltip = "text")
```



## Dashboard

```{r dashboard}
# UI layout using shinydashboard
ui <- dashboardPage(
    dashboardHeader(title = "Malaria Data Visualization Dashboard"),
    dashboardSidebar(
        sidebarMenu(
            menuItem("Dashboard", tabName = "dashboard", icon = icon("dashboard")),
            menuItem("Help", icon = icon("question-circle"))
        )
    ),
    dashboardBody(
        tabItems(
            tabItem(tabName = "dashboard",
                    fluidRow(
                        box(plotlyOutput("plot1", height = "450px")),
                        box(plotlyOutput("plot2", height = "450px"))
                    )
            )
        )
    )
)

# Server logic
server <- function(input, output) {
    # Load and preprocess your data here

    # Plot 1: Americas
    output$plot1 <- renderPlotly({

# year 2000
latest_2000 <- 2000
merged_death_2000 <- subset(merged_death, Year == latest_2000)
merged_mort_2000 <- subset(merged_mort, Year == latest_2000)
merged_incd_2000 <- subset(merged_incd, Year == latest_2000)


merged_death_wp <- subset(merged_death_2000, Region == 'Western Pacific')
merged_mort_wp <- subset(merged_mort_2000, Region == 'Western Pacific')
merged_incd_wp <- subset(merged_incd_2000, Region == 'Western Pacific')

# Merging the death and mortality data for the WPs
merged_data_wp <- merge(merged_death_wp, merged_mort_wp, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the WP region
p_wp <- ggplot(merged_data_wp, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region, text=Country)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
  scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='West_Pacific Estimates (2000)',
         x='Estimated Deaths in 2000',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right') 

# Converting to an interactive plotly plot
ggplotly(p_wp, tooltip = "text")
})

    # Plot 2: Africa Bubble Plot
    output$plot2 <- renderPlotly({
        
## 2021
latest_2021 <- 2021
merged_death_2021 <- subset(merged_death, Year == latest_2021)
merged_mort_2021 <- subset(merged_mort, Year == latest_2021)
merged_incd_2021 <- subset(merged_incd, Year == latest_2021)


merged_death_wp <- subset(merged_death_2021, Region == 'Western Pacific')
merged_mort_wp <- subset(merged_mort_2021, Region == 'Western Pacific')
merged_incd_wp <- subset(merged_incd_2021, Region == 'Western Pacific')

# Merging the death and mortality data for the Americas
merged_data_wp <- merge(merged_death_wp, merged_mort_wp, by=c('Country', 'Year', 'Region'))

# Creating a ggplot for the WP region
p_wp <- ggplot(merged_data_wp, aes(x=Est_Death, y=Est_Mort, size=Est_Death, color=Region)) +
    geom_point(alpha=0.7) +
    geom_text(aes(label=Country, color = "black"), check_overlap = TRUE, size = 1) +
    scale_size(range=c(1, 10)) +
    scale_color_manual(values=c("Africa" = "red", "Americas" = "skyblue", "Europe" = "red", "South-East Asia" = "orange", "Eastern Mediterranean" = "green", "Western Pacific" = "yellow")) +
    labs(title='West_Pacific Estimates (2021)',
         x='Estimated Deaths in 2021',
         y='Average Estimated Mortality Rate',
         color='Region') +
    theme_classic() +
    theme(legend.position='right')

# Converting to an interactive plotly plot
ggplotly(p_wp, tooltip = "text")
    })
}

# Run the application
shinyApp(ui = ui, server = server)
```

